Natural Selection Across Three Generations of Americans

### 2024-02-19 response-to-reviewers-revision-1.qmd

---
title: "'Natural Selection Across Three Generations of Americans': response to reviewers"
format: docx
editor: 
  markdown: 
    wrap: 72
---

@@

Dear Professor Medland,

Thank you for the opportunity to revise our manuscript. Thank you also
to the reviewers. We respond to their points below.

One major change is that I have asked Tobias Edwards to be a co-author.
I discovered that he was working on similar topics, so we merged our
efforts. One effect is just that we have more
polygenic scores to analyse. The original results are not much changed,
which is encouraging. Another is that we look in
more detail at two extra generations, using children and grandparents of
the respondents. We've given the paper a title which reflects that. Lastly,
we have more careful and accurate estimates of effect sizes of "true" and/or
"chip-wide" scores. I hope this is OK.

@@

*In a similar vein, it would be very welcome to reiterate reason why the
expected effect is stronger amongst low income etc people.*

We now detail the theory more in the abstract, and we've described the
substitution effects and income effects straight away in the intro.
We've tried to keep the language as simple and intuitive as possible.

*I was wondering why the authors control for PCAs in the PGI analyses.
As stated by the authors, this study is not a causal study but one of
correlations. The consequence of a correlation (natural selection)
should be independent of its causes.*

Note that controlling for principal components in a regression, *y \~
PGIs + PCs*, gives the same effect sizes as running an uncontrolled
regression on PGIs which have previously had the PCs residualized out,
*y \~ PGIs_resid*. We control rather than residualizing because this
gives the correct standard errors[^1], which might matter with our
relatively low N.

[^1]: Because the controls are correctly accounted for when calculating
    degrees of freedom.

@@

Negatively correlated, and more strongly, opposite-signed, effects
between young and old parents support the theory that income effects
dominate for older parents while substitution effects dominate for
younger parents. To be a bit more explicit: if a PGS simply doesn't
correlate with human capital, we expect it to not be selected on among
either group of parents. If a PGS is correlated with human capital, that
will lead to income and substitution effects for those with more human
capital. If the substitution effect dominates for younger parents then
the PGS will have a negative correlation with fertility; if the income
effect dominates for older parents, then the PGS will have a positive
correlation with fertility.

The test against 29 had the following logic: if all effects were really
zero, and estimates were pure noise with mean 0, then we'd expect
opposite-signed and same-signed "effects" to be equally likely; so, 29
is half of the 58 scores.

However, thinking about this issue a bit more made us change our tests
more generally. The original situation was that we tested most pairs of
coefficients across sample subgroups by comparing their absolute size.
Coefficients for older and younger parents had similar absolute sizes,
but we noticed they were opposite-signed and hence ran the tests above.
This is not very satisfactory: it's too like special pleading, and it
also suggests the original tests weren't adequate.

We now improve our tests in table 1 as follows. We take the null
hypothesis to be "selection coefficients are the same in the different
subgroups, for all PGIs". Under the null, the true difference between
selection coefficients is zero; the difference in our sample (and in
bootstrapped subsamples) should then not be significantly different from
zero. Our alternative hypothesis is that, for at least some scores,
selection coefficients against human capital are smaller among the
richer/more educated/older/married group. That is, if the selection
coefficient among the poorer and richer groups are $x_p,x_r$, and
$sign(corr)$ is the sign of the PGI's correlation with human capital,
then

$x_r \times -sign(corr) < x_p \times -sign(corr)$

@@

*However, in my opinion the theoretical database, data and analysis are
adequate, the problem I have is that the MS is a bit confusing. It
should be made clearer why an analysis was done and what the results
mean in terms of rejecting or accepting the theoretical assumptions. The
possible link to biological selection should also be explained in more
detail, as this has been done in the case of education, income and
reproduction to some extent on the basis of comparable data (Kong et al.
2017, Beauchamp 2016, Fieder & Huber 2002) and also in many other
studies.*

We have clarified the theory early on in the intro and abstract. TE
TODO: Write a paragraph on biol. selection. (Not sure what is meant
here?)

In terms of accepting or rejecting our theory, our paragraph in the
discussion was:

> The results here support prediction 1 but are more ambiguous for
> prediction 2. PGS coefficients on RLRS also appear larger for
> low-income groups and unmarried respondents. But there is little
> evidence for larger coefficients among people with lower education, or
> younger parents. This may be due to the low sample size. But in the
> UK, the between-group differences were large; differences that big
> would surely have been visible here. The theory can accommodate this
> non-result, if preferences for income are less curved in the US for
> whatever reason. But note that *any* theory with a negative
> relationship between education and fertility will give prediction 1.
> In this sense, results here are less supportive of the economic theory
> specifically.

@@

We don't think these results give a simple thumbs up or down to the
theory, but we're open to suggestions that would help clarify what we
say.

*The division into a "black" and a "white" is necessary due to technical
reasons, but further complicates the results/discussion. Presumably the
results for the larger sample (I think in the case of the HRS this are
the whites) should be presented in the main text, and the results for
the smaller population moved into a supplement?*

Done. Indeed, the black sample is really too small to say much.

@@

*Also, there seem to be many associations of PGS with RLRS (Figure 4),
such as ADHD, diabetes, etc., which I think are interesting findings,
even if of low significance, worth at least a brief discussion.*

We agree. We now discuss some health-related traits in the conclusion:

> Whilst past research has focused on \[cognitive\] traits, we find
> selection differentials are often of similar magnitudes for health
> related traits. The selection differential for self-rated health is
> greater than those of education and cognitive performance. The most
> significant, positively selected trait was ADHD, which was also found
> in the UK [@hugh2022human]. Future research should study the health
> and medical implications of natural selection, in addition to their
> social implications.

@@

*What is the data base for the mean number of children of people born in
the same year - is pooling appropriate, as the means may vary
considerably according to ethnicity?*

We thought about this, and now agree that it is better to use mean number of
children of whites and of blacks separately for each ethnic group:

> "RLRS is calculated separately by ethnicity. This is not ideal, because it treats the ethnicities as separate breeding populations. But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative. We therefore focus on the white US population, with the caveat that this "one data point" may not be representative of the US population (still less the species!) The mean number of children of people born in each year was calculated using sampling weights."

Effectively, the white US population is a "single data
point" for theory testing. Note that since we don't compare the ethnicities,
the only difference this makes is in estimating the absolute size of effects.

TE TODO: change it to be separate RLRS baseline for white and black.

### 2024-02-19 hrs-selection.qmd
Discuss separate vs pooled RLRS. Delete some out-of-date bits.

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year. RLRS is calculated pooling ethnicities, i.e. treating them as members of the same biological population. The mean number of children of people born in the same year was calculated using sampling weights.

The HRS contains weights which match survey respondents to the US population. We use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, we weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012. Statistical tests are adjusted for clustering and stratification using the R "survey" package [@lumleysurvey2023].

@@

```{r}
#| label: fig-deltas
#| fig-cap: Coefficients of interaction terms ($\delta$) for effects of
#|   polygenic scores 
#|   on RLRS among different social groups. A positive $\delta$ means that
#|   the coefficient was larger for the disadvantaged group, where the score
#|   was sign-flipped to correlate negatively with education. Horizontal lines
#|   show medians.
deltas |> 
  filter(
    group != "born"
  ) |> 
  mutate(
    med_d = median(delta),
    Group = case_match(group,
              "agefb"      ~ "AFB",
              "edu"        ~ "Education",
              "income_med" ~ "Income",
              "married"    ~ "Marriage"
            ),
    outliers = ifelse(rank(delta) %in% c(1:3, 66:68), PGS, ""),
    .by = group
  ) |> 
  ggplot(aes(Group, delta, colour = Group)) + 
    geom_hline(yintercept = 0, color='grey70') + 
    geom_point(alpha = 0.7, position = position_jitter(width = 0.1)) +
    geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
    geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
    theme_minimal() + 
    theme(
      panel.grid = element_blank(),
      legend.position = "none"
    )
```

@@

: Estimates and bootstrap 95% confidence intervals for numbers of PGS out of `r length(pgs_white)` where the disadvantaged group had a larger selection coefficient than the advantaged group. Groups are: 0-12 years education vs. 13-17 years; below vs. above median income; all others vs. married; below vs. above median age at first birth; born before vs. after 1942. `r n_reps_x_es` bootstraps.

@@

```{r}
#| label: stat-x-bootstrap-birthyear
#| cache: false

n_reps_birthyear <-  199
pgs <- pgs_white

forms <- pgs |> map(\(x) reformulate(c(x, pcs), response = "raevbrn"))

calc_birthyear <- function (w, data) {
  data$w <- w
  mods_young <- map(forms, \(x) lm(x, data, subset = rabyear > 1945,
                                 weights = w))
  mods_old <- map(forms, \(x) lm(x, data, subset = rabyear <= 1945,
                                  weights = w))
  coefs_young <- map2_dbl(mods_young, pgs, \(x, y) coef(x)[[y]])
  coefs_old <- map2_dbl(mods_old, pgs, \(x, y) coef(x)[[y]])

  c(cor = cor(coefs_young, coefs_old))
}

rand_boot <- rand |> 
  subset(ethnicity == "white") |> 
  as.svrepdesign(type = "bootstrap", replicates = n_reps_birthyear)

cor_agefb <- withReplicates(rand_boot, theta = calc_birthyear)

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority, @johnson1979minority, @yang2003big). And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated. But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative. Looking within the sample, changes in PGS by birth year are small for both groups, and are probably mostly driven by selective mortality. Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s. The last line of @tbl-x-bootstraps shows that effect sizes are larger for respondents born after 1942, but the difference is imprecisely estimated and not significant.

```{r}
#| label: ADHD
#| cache: false

# Testing the R2 of the ADHD PGS
rand$variables |>
  with(cor.test(adhd_score, ADHD, use = "pairwise.complete")) -> adhd_cor

>>>>

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year. RLRS is calculated separately by ethnicity. This is not ideal, because it treats the ethnicities as separate breeding populations. But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative. We therefore focus on the white US population, with the caveat that this "one data point" may not be representative of the US population (still less the species!) The mean number of children of people born in each year was calculated using sampling weights.

The HRS contains weights which match survey respondents to the US population. We use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, we weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012. Statistical tests are adjusted for clustering and stratification using the R *survey* package [@lumleysurvey2023].

@@

```{r}
#| label: fig-deltas
#| fig-cap: Coefficients of interaction terms ($\delta$) for effects of
#|   polygenic scores 
#|   on RLRS among different social groups. A positive $\delta$ means that
#|   the coefficient was larger for the disadvantaged group, where the score
#|   was sign-flipped to correlate negatively with education. Horizontal lines
#|   show medians.

deltas |> 
  filter(
    group != "born"
  ) |> 
  mutate(
    med_d = median(delta),
    Group = case_match(group,
              "agefb"      ~ "AFB",
              "edu"        ~ "Education",
              "income_med" ~ "Income",
              "married"    ~ "Marriage"
            ),
    outliers = ifelse(delta > 0.05, nice_names[PGS], ""),
    .by = group
  ) |> 
  ggplot(aes(Group, delta, colour = Group)) + 
    geom_hline(yintercept = 0, color='grey70') + 
    geom_point(alpha = 0.7, 
               position = position_jitter(width = 0.1, seed = 12345)) +
    geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
    geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
    geom_text(aes(label = outliers), size = 2, colour = "black", 
              position = position_jitter(width = 0.1, seed = 12345), 
              hjust = -0.2) +
    theme_minimal() + 
    theme(
      panel.grid = element_blank(),
      legend.position = "none"
    )
```

@@

: Estimates and bootstrap 95% confidence intervals for numbers of PGS out of `r length(pgs_white)` where the disadvantaged group had a larger selection coefficient than the advantaged group. Groups are: 0-12 years education vs. 13-17 years; below vs. above median income; all others vs. married; below vs. above median age at first birth; born before vs. after 1942. `r n_reps_x_es` bootstraps. {#tbl-x-bootstraps}

@@

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority, @johnson1979minority, @yang2003big). And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated. But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative. Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.

```{r}
#| label: ADHD

# Testing the R2 of the ADHD PGS

adhd_cor <- cor.test(rand$variables$adhd_score, rand$variables$ADHD, 
                     use = "pairwise.complete")
